Conceptualizing open health data platforms for low- and middle income countries

### 2024-05-27 index.qmd
wip - rewrite research questions

Conceptually, the OpenHIE framework constitutes a framework for an open digital platform. Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods [@digitalpublicgoods]. If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse. The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.  The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open HDP can be realized that supports the four types of data sharing as defined by de Reuver at al. (@tbl-types-data-sharing).

|   | __Type of secondary data sharing__ | __Technology enablers to create openness of health data platforms__ |
|:-:|:-------|:-------------|
| __1__ | Data at the most granular (patient) level, which is persisted and used to provide a longitudinal record. | __Bulk FHIR API__: has by now been incorporated in all major FHIR implementations the FHIR standard can be readily used to support data sharing to patient-level data across a patient population [@mandl2020push;@jones2021landscape]. |
| __2__ | Aggregated data, for example statistics for policy evaluation and benchmarking |__SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows. Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items. Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].|
| __3__ | Data analytics modules, that provide access to work and access the data. | __Federated learning (FL)__ [@rieke2020future] and __privacy- enhancing technologies (PETs)__ [@scheibner2021revolutionizing;@jordan2022selecting]: new paradigms that address the problem of data governance and privacy by training algorithms collaboratively without exchanging the data itself. Models can be trained on combined datasets and made available as open source artifacts for decision support. Data analysts can use FL and PETs to work with the data in a collaborative, decentralized fashion. |
| __4__ | Trained models that have been derived from the data and can be used stand-alone for decision support. | __ONNX__ [@onnx]: an open format built to represent machine learning models. |

: Types of data sharing and in relation to new standards and technology enablers to create openness. Taken from de Reuver et al. (2022). {#tbl-types-data-sharing}

@@

>>>>

Conceptually, the OpenHIE framework constitutes a framework for an open digital platform. Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods [@digitalpublicgoods]. If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse. The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.  The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open HDP can be realized that supports the following different types of data sharing.

#### 1. Sharing of data sets at the most granular (patient) level, persisted as longitudinal records
The Shared Health Record (SHR) in OpenHIE provides an operational, real-time transactional data source which is intended for primary data sharing. The specification explicitly states that the SHR is distinct from a datawarehouse. At the same time, the Health Management Information System (HMIS) component is specified to fulfil functional requirements that are typically provided by datawarehouses.

- **Q1:** How can proven solution designs and modern technologies from data warehousing and engineering be integrated in the OpenHIE specification and how does this relate to the curren specifications of the SHR and HMIS.

#### 2. Sharing of data products derived from the original data set
OpenHIE specifies that the Health Management Information System (HMIS) component should support a workflow to validate and save aggregate data based on the emerging [IHE Aggregate Data Exchange (ADX) standard](https://www.ihe.net/uploadedFiles/Documents/QRPH/IHE_QRPH_Suppl_ADX.pdf). This is just one specific workflow within a larger data analytics value chain that covers i) collection, ii) standardization, iii) cleaning, iv) storage, v) analysis and vi) distribution [@curry2016big]. Sharing of benchmarking, and decision-support tools based on various statistical (machine learning) models that have been trained on data sets and/or data product, can be shared for standalone use.

- **Q2:** How can proven solutions designs from the data warehousing and engineering community be included in OpenHIE to support sharing of wider variety of data products.

#### 3. Sharing of secure computational environments to access and work with the data

Current solution design of centralized data storage is increasingly being challenged. Recent work points to federated learning (FL) [@rieke2020future] and privacy- enhancing technologies (PETs) [@scheibner2021revolutionizing;@jordan2022selecting] as a better long-term solution to secure and equitable sharing of data. Given the resource constrained environments of LMICs, is is particularly challenging to devise a solution that is economically feasible and ... [TO DO: add feedback from Mark vdG]

Federated solutions:

- KETOS OMOP-FHIR [@gruendner2019ketos]
- Personal Health Train on FHIR [@choudhury2020personal]
- OHDSI analytics [@khalid2021standardized]
- CODA project [@mullie2023coda]
- GenoMed4All, in evalution phase for -omis [@cremonesi2023need]

Similar approaches taken in the context of data sharing for pandemic response, for example:

- VODAN Africa, based on electronic data capture with templates [@purnamajati2022data]
- Global Data Sharing Initiative (GDSI) [@pirmani2023journey]

- **Q3:** How can a solution be designed that is economically viable, can be deployed in a bottom-up fashion, support data governance principles ...

@@

TO DO: re-use these elements from table 1 into this section:

- Given that the Bulk FHIR API has by now been incorporated in all major FHIR implementations [@mandl2020push;@jones2021landscape], the FHIR standard can be readily used to bridge the gap between primary (transactional) and secondary (analytics) data sharing.
- __SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows. Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items. Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].  Modern data platforms take this Within the HL7 ecosystem much progress has been made to support "analytics-on-FHIR" with standards including [FHIRPath](https://hl7.org/fhirpath/) and the new SQL-on-FHIR specification [@sql-on-fhir]. Given the use of FHIR as the common data model (the rationale of which will be described later), we formulate the second design question:

### 2024-05-27 index.tex
wip - rewrite research questions

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Paula van Brakel; Mark van der Graaf; Nicole Spieker; Mark de Reuver},
  pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
platforms, LMICs, digital health},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\author[4]{Antragama Abbas%
%
}

\author[5]{Steven Wanyee%
%
}

\author[6]{Alessandro Pietrobon%
%
}

\author[7]{Ryan Chrichton%
%
}

\author[8]{John Grimes%
%
}

@@

\author[4]{Mark de Reuver%
%
}

\affiliation[1]{organization={PharmAccess
Foundation},city={Amsterdam},country={the
Netherlands},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Eindhoven University of
Technology},city={Eindhoven},country={the
Netherlands},countrysep={,},postcodesep={}}
\affiliation[3]{organization={PharmAccess
Kenya},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[4]{organization={Delft University of
Technology},city={Delft},country={the
Netherlands},countrysep={,},postcodesep={}}
\affiliation[5]{organization={IntelliSOFT},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[6]{organization={ONA},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[7]{organization={Jembi Health
Systems},city={Durban},country={South
Africa},countrysep={,},postcodesep={}}
\affiliation[8]{organization={Australian e-Health Research
Centre},city={Brisbane},country={Australia},countrysep={,},postcodesep={}}

@@

Case studies on digital platforms in healthcare point to an emerging
pattern where the focus shifts from the digital platform with its
defining software and hardware components, to the data as the primary
object of interest in and of itself
\citep{ozalp2022digital, alaimo2022organizations}. This development
roughly correlates with the commonly used distinction between primary
and secondary health data sharing \citep{cascini2024health}. While
people are generally willing to share their health data to receive the
best care (primary use), the attitude towards secondary use of health
data varies greatly depending on the type and context. The shift in
perspective from digital platforms supporting primary data sharing
toward data platforms supporting secondary data sharing, ties in with
the proposed research agenda by de Reuver et al.~to consider data
platforms as a phenomenon distinct from digital platforms
\citep{dereuver2022openness}. Generally, data platforms inherit the
characteristics of digital platforms. From an economic perspective, for
example, both types exhibit multi-sided markets. At the same time, data
platforms differ as their main offerings revolve around data. From an
ecosystem perspective, data platforms have more moderate network effects
and are more susceptible to fragmentation and heterogeneity.

Particularly relevant in the context of health data platforms (HDPs), is
the conceptualization of openness. The shift in perspective from digital
platforms to data platforms coincides with the paradox of open
\citep{keller2021paradox}. Originally, openness of digital platforms
focused on open source, open standards and copyrights, which by has been
superseded by ``\ldots{} conflicts about privacy, economic value
extraction, the emergence of artificial intelligence, and the
destabilizing effects of dominant platforms on (democratic) societies.
Instead of access to information, the control of personal data has
emerged in the age of platforms as the critical contention.''
\citep{keller2021paradox}. These conflicts are particularly salient in
the healthcare domain, as is evidenced by the polemic surrounding data
spaces \citep{otto2022designing} and data solidarity
\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond}.
Openness is particularly relevant if we are to realize a
solidarity-based approach to health data sharing that i) gives people a
greater control over their data as active decision makers; ii) ensures
that the value of data is harnessed for public good; and iii) moves
society towards equity and justice by counteracting dynamics of data
extraction \citep{prainsack2022data}.

@@

This paper is motivated by the conflation of a number of developments
relevant to the design and implementation of open, solidarity-based HDPs
in LMICs. First, the OpenHIE framework \citep{openhie} has been adopted
by many sub-Saharan African countries \citep{mamuye2022health} as the
architectural blueprint for implementing nation-wide health information
exchanges (HIE), including Nigeria \citep{dalhatu2023paper}, Kenya
\citep{thaiya2021adoption} and Tanzania \citep{nsaghurwe2021one}. These
countries have, as a matter of course, extended the framework to include
``analytics services'' as an additional domain. The rationale for this
addition is to facilitate secondary reuse of health data for academic
research, real-world evidence studies etc. which can be framed within
the context of ongoing efforts towards Findable, Accessible,
Interoperable and Reusable (FAIR) sharing of health data
\citep{guillot2023fair}. In doing so, however, we have implicitly moved
from conceptualizing digital health platforms (the original OpenHIE
specification) to health data platforms. This is problematic because the
notion of openness, which is assumed to be essential in establishing
solidarity-based approaches to data sharing, is inherently different for
a data platform compared to a digital platform.

Conceptually, the OpenHIE framework constitutes a framework for an open
digital platform. Openness for digital platforms refers to i) the use of
open boundary resources, that is, specifications for the various
healthcare specific workflows and information standards such as FHIR;
and ii) the use of open source components that are available as digital
public goods \citep{digitalpublicgoods}. Typically, these open source
components are intended to support deployment in small countries
(population up to 10 million) or large NGOs out of the box, and should
provide a stepping stone for customized deployments in medium-sized
countries (population around 40 million).\footnote{Although the OpenHIE
  specification does not include details on dimensioning, these are
  typically the requirements that are used within the community. See
  \href{https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements}{OpenHIE
  Community Wiki}.} If we are to use the OpenHIE framework as an open
data platform, we need to extend the standards, technologies and
architecture to include functionality for data sharing and reuse. The
lack of detailed specifications and consensus of this addition to
OpenHIE currently stands in the way of development projects that aim to
establish HDPs in LMICs. The purpose of this paper, therefore, is to
specificy how new standards and technologies can be integrated into the
OpenHIE architecture such that an open HDP can be realized that supports
the four types of data sharing as defined by de Reuver at
al.~(Table~\ref{tbl-types-data-sharing}).

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1200}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3200}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5600}}@{}}
\caption{Types of data sharing and in relation to new standards and
technology enablers to create openness. Taken from de Reuver et
al.~(2022).}\label{tbl-types-data-sharing}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Type of data sharing}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Technology enablers to create openness of health data platforms}
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Type of data sharing}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Technology enablers to create openness of health data platforms}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{1} & Data at the most granular (patient) level, which is
persisted and used to provide a longitudinal record. & \textbf{Bulk FHIR
API}: has by now been incorporated in all major FHIR implementations the
FHIR standard can be readily used to support data sharing to
patient-level data across a patient population
\citep{mandl2020push, jones2021landscape}. \\
\textbf{2} & Aggregated data, for example statistics for policy
evaluation and benchmarking & \textbf{SQL-on-FHIR specification}
\citep{sql-on-fhir}: provides a standardised approach to make FHIR work
well with familiar and efficient SQL engines that are most commonly used
in analytical workflows. Builds on FHIRPath \citep{hl72020fhirpath}
expressions in a logical structure to specify things like column names
and unnested items. Implementations of this approach are available or
forthcoming, including open source implementations such as Pathling
\citep{grimes2022pathling} and commercial offerings like Aidbox
\citep{aidbox}. \\
\textbf{3} & Data analytics modules, that provide access to work and
access the data. & \textbf{Federated learning (FL)}
\citep{rieke2020future} and \textbf{privacy- enhancing technologies
(PETs)} \citep{scheibner2021revolutionizing, jordan2022selecting}: new
paradigms that address the problem of data governance and privacy by
training algorithms collaboratively without exchanging the data itself.
Models can be trained on combined datasets and made available as open
source artifacts for decision support. Data analysts can use FL and PETs
to work with the data in a collaborative, decentralized fashion. \\
\textbf{4} & Trained models that have been derived from the data and can
be used stand-alone for decision support. & \textbf{ONNX} \citep{onnx}:
an open format built to represent machine learning models. \\
\end{longtable}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  the OpenHIM platform
  (\url{https://jembi.gitbook.io/openhim-platform/})
\item
  the ONA Canopy platform \url{https://ona.io/home/services/canopy/}
\item
  the work conducted at PharmAccess Foundation as part of the MomCare
  programme (\url{https://health-data-commons.pharmaccess.org}).
\end{enumerate}

@@

@@

While the lakehouse architecture separates the concerns of compute and
storage, the composable data stack takes the separation of concerns is
taken one step further. A composable data system
(Figure~\ref{fig-composable-data-stack}), not only separates the storage
(layer 3) and execution (layer 2), but also separates the user interface
(layer 1) from the execution engine by introducing standards including
Substrait for Intermediate Representation (standard A, IR) and Apache
Arrow for data connectivity and data memory layout (standards B and C,
respectively). The composable data stack can be implemented with current
open source technologies (\textbf{?@fig-cds-examples}). As an example,
the Ibis user interface is currently sufficiently mature to offer a
standardized dataframe interface to 19 different execution engines.

@@

Today, many components of the OpenHIE specification are now available as
a digital public goods, as listed in
Table~\ref{tbl-digital-public-goods}. To further ease the development,
configuration and deployment of health information exchanges, the
concept of `Instant OpenHIE' has been championed to (i) allow
implementers to engage with a preconfigured health information exchange
solution and running tools (based on the architecture) and test their
applicability and functionality with a real health context problem; and
(ii) have a packaged reference version of the OpenHIE architecture that
is comprised of a set of reference technologies and other appropriate
tools that form the building blocks of the health information exchange
that can be configured and extended to support particular use cases
\citep{InstantOpenHIEv2}. Besides the core functional components of the
OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and
integration of generic components such as Identity and Access Managment
(IAM) and a reverse proxy gateway. In the following, we will evaluatie
three of such configurations, with the aim to conceptualize and evaluate
the proposed Data and Analytics Services domain of of the OpenHIE
architecture.

@@

Clearly, a more open and scaleable platform was required if MomCare was
to be implemented in more regions. This need led to a redesign of the
underlying technical infrastructure of the MomCare project. The
objectives of this work were in fact to demonstrate a solution design
that could support the first three types of data sharing listed in
Table~\ref{tbl-types-data-sharing}. First, to investigate the viability
of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed
to assess the complexity and effort required to implement the facade
pattern to integrate the legacy system into the FHIR data standard.
Using the longitudinal dataset from approximately 28 thousand patient
records, FHIR transformations script were developed and deployed using
the mediator function of the IOL. The data was transformed into 10 FHIR
v4 resources and the conceptual data model of the existing MomCare app
could readily be transformed into the FHIR standard using SQL and
validated with a Python library \citep{islam2023fhir}. The largest
challenge during the transformation process pertained to the absence of
unique business identifiers for patients and healthcare organizations.
For patients, either the mobile phone number or the healthcare insurance
number was taken, depending on availability. A combination of name,
address and latitude/longitude coordinates were used to uniquely
identify organizations and locations, as Tanzania does not have a system
in place for this purpose.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Paula van Brakel; Mark van der Graaf; Nicole Spieker; Antragama Abbas; Mark de Reuver},
  pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
platforms, LMICs, digital health},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\author[4]{Steven Wanyee%
%
}

\author[5]{Alessandro Pietrobon%
%
}

\author[6]{Ryan Chrichton%
%
}

\author[7]{John Grimes%
%
}

@@

\author[8]{Antragama Abbas%
%
}

\author[8]{Mark de Reuver%
%
}

\affiliation[1]{organization={PharmAccess
Foundation},city={Amsterdam},country={the
Netherlands},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Eindhoven University of
Technology},city={Eindhoven},country={the
Netherlands},countrysep={,},postcodesep={}}
\affiliation[3]{organization={PharmAccess
Kenya},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[4]{organization={IntelliSOFT},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[5]{organization={ONA},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
\affiliation[6]{organization={Jembi Health
Systems},city={Durban},country={South
Africa},countrysep={,},postcodesep={}}
\affiliation[7]{organization={Australian e-Health Research
Centre},city={Brisbane},country={Australia},countrysep={,},postcodesep={}}
\affiliation[8]{organization={Delft University of
Technology},city={Delft},country={the
Netherlands},countrysep={,},postcodesep={}}

@@

Case studies on digital platforms in healthcare point to an emerging
pattern where the focus shifts from the digital platform with its
defining software and hardware components, to the data as the primary
object of interest in and of itself
\citep{ozalp2022digital, alaimo2022organizations}. This observation ties
into with the proposed research agenda by de Reuver et al.~to consider
data platforms as a phenomenon distinct from digital platforms
\citep{dereuver2022openness}. Generally, data platforms inherit the
characteristics of digital platforms. From an economic perspective, for
example, both types exhibit multi-sided markets. At the same time, data
platforms differ as their main offerings revolve around data. From an
ecosystem perspective, data platforms have more moderate network effects
and are more susceptible to fragmentation and heterogeneity.

Particularly relevant in the context of health data platforms (HDPs), is
the conceptualization of openness. The shift in perspective from digital
platforms to data platforms coincides with the paradox of open
\citep{keller2021paradox}. Originally, openness of digital platforms
focused on open source, open standards and copyrights, which by has been
superseded by ``\ldots{} conflicts about privacy, economic value
extraction, the emergence of artificial intelligence, and the
destabilizing effects of dominant platforms on (democratic) societies.
Instead of access to information, the control of personal data has
emerged in the age of platforms as the critical contention.''
\citep{keller2021paradox}. These conflicts are particularly salient in
the healthcare domain, where people are generally willing to share their
health data to receive the best care (primary use), while the attitude
towards secondary use of health data varies greatly depending on the
type and context \citep{cascini2024health}. The shift in perspective
from digital platforms supporting primary data sharing toward data
platforms supporting secondary data sharing is one of the key issues
surrounding the polemic of data spaces \citep{otto2022designing} and
data solidarity
\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond}.
Openness is particularly relevant if we are to realize a
solidarity-based approach to health data sharing that i) gives people a
greater control over their data as active decision makers; ii) ensures
that the value of data is harnessed for public good; and iii) moves
society towards equity and justice by counteracting dynamics of data
extraction \citep{prainsack2022data}.

@@

This paper is motivated by the conflation of a number of developments
relevant to the design and implementation of open, solidarity-based HDPs
in LMICs. First, the OpenHIE framework \citep{openhie} has been adopted
by many sub-Saharan African countries \citep{mamuye2022health} as the
architectural blueprint for implementing nation-wide health information
exchanges (HIE), including Nigeria \citep{dalhatu2023paper}, Kenya
\citep{thaiya2021adoption} and Tanzania \citep{nsaghurwe2021one}. These
countries have, as a matter of course, extended the framework to include
``data \& analytics services'' as an additional domain. The rationale
for this addition is to facilitate secondary reuse of health data for
academic research, real-world evidence studies etc. which can be framed
within the context of ongoing efforts towards Findable, Accessible,
Interoperable and Reusable (FAIR) sharing of health data
\citep{guillot2023fair}. In doing so, however, we have implicitly moved
from conceptualizing digital health platforms for primary data sharing
(the original OpenHIE specification) to health data platforms for
secondary data sharing. This is problematic because the notion of
openness, which is assumed to be essential in establishing
solidarity-based approaches to data sharing, is inherently different for
a data platform compared to a digital platform.

Conceptually, the OpenHIE framework constitutes a framework for an open
digital platform. Openness for digital platforms refers to i) the use of
open boundary resources, that is, specifications for the various
healthcare specific workflows and information standards such as FHIR;
and ii) the use of open source components that are available as digital
public goods \citep{digitalpublicgoods}. If we are to use the OpenHIE
framework as an open data platform, we need to extend the standards,
technologies and architecture to include functionality for secondary
data sharing and reuse. The lack of detailed specifications and
consensus of this addition to OpenHIE currently stands in the way of
development projects that aim to establish HDPs in LMICs. The purpose of
this paper, therefore, is to specificy how new standards and
technologies can be integrated into the OpenHIE architecture such that
an open HDP can be realized that supports the following different types
of data sharing.

\subsubsection{1. Sharing of data sets at the most granular (patient)
level, persisted as longitudinal
records}\label{sharing-of-data-sets-at-the-most-granular-patient-level-persisted-as-longitudinal-records}

The Shared Health Record (SHR) in OpenHIE provides an operational,
real-time transactional data source which is intended for primary data
sharing. The specification explicitly states that the SHR is distinct
from a datawarehouse. At the same time, the Health Management
Information System (HMIS) component is specified to fulfil functional
requirements that are typically provided by datawarehouses.

\begin{itemize}
\tightlist
\item
  \textbf{Q1:} How can proven solution designs and modern technologies
  from data warehousing and engineering be integrated in the OpenHIE
  specification and how does this relate to the curren specifications of
  the SHR and HMIS.
\end{itemize}

\subsubsection{2. Sharing of data products derived from the original
data
set}\label{sharing-of-data-products-derived-from-the-original-data-set}

OpenHIE specifies that the Health Management Information System (HMIS)
component should support a workflow to validate and save aggregate data
based on the emerging
\href{https://www.ihe.net/uploadedFiles/Documents/QRPH/IHE_QRPH_Suppl_ADX.pdf}{IHE
Aggregate Data Exchange (ADX) standard}. This is just one specific
workflow within a larger data analytics value chain that covers i)
collection, ii) standardization, iii) cleaning, iv) storage, v) analysis
and vi) distribution \citep{curry2016big}. Sharing of benchmarking, and
decision-support tools based on various statistical (machine learning)
models that have been trained on data sets and/or data product, can be
shared for standalone use.

\begin{itemize}
\tightlist
\item
  \textbf{Q2:} How can proven solutions designs from the data
  warehousing and engineering community be included in OpenHIE to
  support sharing of wider variety of data products.
\end{itemize}

\subsubsection{3. Sharing of secure computational environments to access
and work with the
data}\label{sharing-of-secure-computational-environments-to-access-and-work-with-the-data}

Current solution design of centralized data storage is increasingly
being challenged. Recent work points to federated learning (FL)
\citep{rieke2020future} and privacy- enhancing technologies (PETs)
\citep{scheibner2021revolutionizing, jordan2022selecting} as a better
long-term solution to secure and equitable sharing of data. Given the
resource constrained environments of LMICs, is is particularly
challenging to devise a solution that is economically feasible and
\ldots{} {[}TO DO: add feedback from Mark vdG{]}

Federated solutions:

\begin{itemize}
\tightlist
\item
  KETOS OMOP-FHIR \citep{gruendner2019ketos}
\item
  Personal Health Train on FHIR \citep{choudhury2020personal}
\item
  OHDSI analytics \citep{khalid2021standardized}
\item
  CODA project \citep{mullie2023coda}
\item
  GenoMed4All, in evalution phase for -omis \citep{cremonesi2023need}
\end{itemize}

Similar approaches taken in the context of data sharing for pandemic
response, for example:

\begin{itemize}
\item
  VODAN Africa, based on electronic data capture with templates
  \citep{purnamajati2022data}
\item
  Global Data Sharing Initiative (GDSI) \citep{pirmani2023journey}
\item
  \textbf{Q3:} How can a solution be designed that is economically
  viable, can be deployed in a bottom-up fashion, support data
  governance principles \ldots{}
\end{itemize}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  the OpenHIM platform
  (\url{https://jembi.gitbook.io/openhim-platform/})
\item
  the OnaData platform
  \url{https://ona.io/home/products/ona-data/features/}
\item
  the work conducted at PharmAccess Foundation as part of the MomCare
  programme (\url{https://health-data-commons.pharmaccess.org}).
\end{enumerate}

@@

TO DO: re-use these elements from table 1 into this section:

\begin{itemize}
\tightlist
\item
  Given that the Bulk FHIR API has by now been incorporated in all major
  FHIR implementations \citep{mandl2020push, jones2021landscape}, the
  FHIR standard can be readily used to bridge the gap between primary
  (transactional) and secondary (analytics) data sharing.
\item
  \textbf{SQL-on-FHIR specification} \citep{sql-on-fhir}: provides a
  standardised approach to make FHIR work well with familiar and
  efficient SQL engines that are most commonly used in analytical
  workflows. Builds on FHIRPath \citep{hl72020fhirpath} expressions in a
  logical structure to specify things like column names and unnested
  items. Implementations of this approach are available or forthcoming,
  including open source implementations such as Pathling
  \citep{grimes2022pathling} and commercial offerings like Aidbox
  \citep{aidbox}. Modern data platforms take this Within the HL7
  ecosystem much progress has been made to support ``analytics-on-FHIR''
  with standards including \href{https://hl7.org/fhirpath/}{FHIRPath}
  and the new SQL-on-FHIR specification \citep{sql-on-fhir}. Given the
  use of FHIR as the common data model (the rationale of which will be
  described later), we formulate the second design question:
\end{itemize}

@@

While the lakehouse architecture separates the concerns of compute and
storage, the composable data stack takes the separation of concerns is
taken one step further. A composable data system
(Figure~\ref{fig-composable-data-stack}), not only separates the storage
(layer 3) and execution (layer 2), but also separates the user interface
(layer 1) from the execution engine by introducing standards including
Substrait for Intermediate Representation (standard A, IR) and Apache
Arrow for data connectivity and data memory layout (standards B and C,
respectively). The first generation of open source components are
already available. For example, the Ibis user interface is currently
sufficiently mature to offer a standardized dataframe interface to 19
different execution engines \citep{ibisproject}.

@@

Today, many components of the OpenHIE specification are now available as
a digital public goods, as listed in
Table~\ref{tbl-digital-public-goods}. Typically, these open source
components are intended to support deployments in small countries
(population up to 10 million) or large NGOs out of the box, and should
provide a stepping stone for customized deployments in medium-sized
countries (population around 40 million).\footnote{Although the OpenHIE
  specification does not include details on dimensioning, these are
  typically the requirements that are used within the community. See
  \href{https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements}{OpenHIE
  Community Wiki}.} To further ease the development, configuration and
deployment of health information exchanges, the concept of `Instant
OpenHIE' has been championed to (i) allow implementers to engage with a
preconfigured health information exchange solution and running tools
(based on the architecture) and test their applicability and
functionality with a real health context problem; and (ii) have a
packaged reference version of the OpenHIE architecture that is comprised
of a set of reference technologies and other appropriate tools that form
the building blocks of the health information exchange that can be
configured and extended to support particular use cases
\citep{InstantOpenHIEv2}. Besides the core functional components of the
OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and
integration of generic components such as Identity and Access Managment
(IAM) and a reverse proxy gateway. In the following, we will evaluatie
three of such configurations, with the aim to conceptualize and evaluate
the proposed Data and Analytics Services domain of of the OpenHIE
architecture.

@@

Clearly, a more open and scaleable platform was required if MomCare was
to be implemented in more regions. This need led to a redesign of the
underlying technical infrastructure of the MomCare project. The
objectives of this work were in fact to demonstrate a solution design
that could support the first three types of data sharing listed in
\textbf{?@tbl-types-data-sharing}. First, to investigate the viability
of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed
to assess the complexity and effort required to implement the facade
pattern to integrate the legacy system into the FHIR data standard.
Using the longitudinal dataset from approximately 28 thousand patient
records, FHIR transformations script were developed and deployed using
the mediator function of the IOL. The data was transformed into 10 FHIR
v4 resources and the conceptual data model of the existing MomCare app
could readily be transformed into the FHIR standard using SQL and
validated with a Python library \citep{islam2023fhir}. The largest
challenge during the transformation process pertained to the absence of
unique business identifiers for patients and healthcare organizations.
For patients, either the mobile phone number or the healthcare insurance
number was taken, depending on availability. A combination of name,
address and latitude/longitude coordinates were used to uniquely
identify organizations and locations, as Tanzania does not have a system
in place for this purpose.
